---
title: "R Notebook"
output: html_notebook
---

```{r}
# don't need to run this code again
install.packages("tidyverse")
```

We load libraries at the beginning of every script

```{r}
library(tidyverse)
```

Load some data

```{r}
weather <- read_csv("data/weather_events.csv")
```

Main tidyverse functions

```{r}
# SORTING FUNCTION
weather %>% arrange(desc(date))

weather %>% arrange(desc(injuries))

# FILTERING 
weather %>% filter(type== "tornado") %>% arrange(desc(fatalities))

weather %>% filter(fatalities > 0)

# operators
# ==
# !=
# >, <, >=, <=
# AND, OR: &, |

weather %>% filter(type == "tornado" & fatalities > 0)

weather %>% filter(type == "tornado" | type == "wind")

# GROUPING
weather %>% group_by(type) %>% summarise(total_injuries = sum(injuries)) %>% arrange(desc(total_injuries))
```

**Homework 1/25:**

1.  What was the most common weather event type in your home state? *or* Which weather event happened most in Illinois?

    Wind was the most common weather event in Illinois in this data set. 

```{r}
weather %>% filter(state == "IL") %>% 
  group_by (type) %>% 
  summarise(number_of_events = n()) %>%
  arrange(desc(number_of_events))
  
```

2.  How many injuries and fatalities were associated with it? *or* How many injuries did Illinois have regarding wind weather events?

    There were 659 total combined injuries and fatalities listed in the data from wind events in Illinois.

```{r}
#ATTEMPTS
# weather %>% filter(state == "IL" & type == "wind") %>% summarise(numbers_of_events = n())

# weather %>% filter(state == "IL" & type == "wind") %>%  group_by (type) %>% summarise(total_injfat = sum(injuries & fatalities))

weather %>% 
  filter(state == "IL" & type == "wind") %>% 
  group_by (type) %>% 
  summarise(total_injfat = sum(injuries, fatalities)) 

###LIZ: Great work problem sovling. Just one small note: I would separate out injuries and fatalities:
weather %>% 
  filter(state == "IL" & type == "wind") %>% 
  group_by(type) %>% 
  summarise(total_inj = sum(injuries), total_fat = sum(fatalities)) 

```

```{r}
#CHECK WORK
#SUM OF INJURIES IN IL FROM WIND = 628
weather %>% 
  filter(state == "IL" & type == "wind") %>% 
  group_by (type) %>% 
  summarise(total_inj = sum(injuries)) 

#SUM OF FATALITIES IN IL FROM WIND = 31
weather %>% 
  filter(state == "IL" & type == "wind") %>% 
  group_by (type) %>% 
  summarise(total_fat = sum(fatalities)) 
```

3.  Which weather event type has the highest fatality rate?

    The weather event type with the highest number of fatalities was tornado with 6,008 fatalities.

```{r}
weather %>% 
  group_by(type) %>% 
  summarise(total_fat = sum(fatalities)) %>% 
  arrange(desc(total_fat))

###LIZ: something to keep in mind here: when comparing the fatalties of different events, you have to take into consideration that some events happen more frequently than others. One way to account for this is to use average: 
weather %>% 
  group_by(type) %>% 
  summarise(avg_fat = mean(fatalities)) %>% 
  arrange(desc(avg_fat))

### same answer, but better process.
```

```{r}
# CHECKING WORK
weather %>% filter(type == "tornado") %>% summarise(total_fat = sum(fatalities))

weather %>% filter(type == "wind") %>% summarise(total_fat = sum(fatalities))

weather %>% filter(type == "hail") %>% summarise(total_fat = sum(fatalities))

###LIZ: love that you're checking your work here!
```

4.  Which county has the most weather events? (use `cnty1_fips`)

    County 31 had the most weather events with 3,011,275,773 weather events present in the data.

```{r}
weather %>% 
  group_by (cnty1_fips) %>% 
  summarise(total = sum(event_num)) %>% 
  arrange(desc(total))

###LIZ: A couple things: an answer that includes 3 billion weather events should trigger a red flag in your brain: if one row is one event, there are only 900K rows. Instead of summing event_num (which is just a random identifier), you want to count rows using n(). Also this was a little bit of a trick question, I realize. County fips codes (three-digit codes) can be repeated in different states, so you really need to use the state column also: 
weather %>% 
  group_by(state, cnty1_fips) %>% 
  summarise(number_of_events=n()) %>% 
  arrange(desc(number_of_events))

###LIZ: then you get a county in Kansas.
```

```{r}
#CHECK WORK
weather %>% 
  group_by (cnty1_fips) %>% 
  summarise(total = sum(event_num)) %>% 
  arrange(total)
```

5.  Are weather events increasing over time?

    Yes, weather events are increasing over time with few years averaging outside of the gradual increase.

```{r}
# ATTEMPT
# weather %>% group_by(year) %>% summarise (avg = avg(event_num))

# plot(weather %>% group_by (year) %>% summarise(avg = mean(event_num))) 

weather %>%
  group_by (year) %>% 
  summarise(avg = mean(event_num)) 

###LIZ: same issue here, just with what you're calculating. Event_num is a meaningless number, you want to count rows: 
weather %>%
  group_by (year) %>% 
  summarise(avg = n())

###LIZ: there's not a very obvious trend. 
```

```{r}
#CHECK
weather %>%
  group_by (year) %>% 
  summarise(avg = mean(event_num)) %>% 
  arrange(desc(avg))
```
