---
title: "Cleaning practice"
output: 
---

LIBRARIES:

```{r}
library(tidyverse)
library(janitor)
library(lubridate)
library(readxl)
```

DATA (located in Files on Canvas):

```{r}
# transactions file from our earlier work with FEC Campaign Finance
transactions <- read_csv("data/transactions.csv")

# deaths file h/t Mary Jo Webster, data editor of the Minnesota Star Tribune
mn_deaths <- read_csv("data/MN_opiate_deaths.csv")

# a Census Bureau file of population by state for five years, 2018 - 2022
state_pop <- read_csv("data/state_pop.csv")
```

TASKS: `transactions`

1.  convert the transaction_dt to a true date column
```{r}
glimpse(transactions)

transactions %>% mutate(trans_dt_proper = mdy(transaction_dt)) %>% 
  select(trans_dt_proper, transaction_dt)

transactions <- transactions %>% mutate(trans_dt_proper = mdy(transaction_dt), .after= transaction_dt)
```

2.  create a new column in the `transactions` table that is a 5-digit zip code (always the first 5 when there are 9 digits.)
```{r}
transactions %>% mutate(new_zipcode = str_sub(zip_code, 1, 5)) %>% 
  select(zip_code, new_zipcode)

transactions <- transactions %>% mutate(new_zipcode = str_sub(zip_code, 1, 5), .after= zip_code)
```


TASKS: `mn_deaths`

1.  check if there are duplicates (using StateID & using names/dob)
There are no duplicates. 
```{r}
glimpse(mn_deaths) #3807 rows

mn_deaths %>% distinct() #3807 rows

mn_deaths %>% count(STATEID) %>% arrange(desc(n))
#no duplicate state id's

mn_deaths %>% count(BIRTHDATE) %>% arrange(desc(n))
# duplicate birthdays

mn_deaths %>% count(BIRTHDATE) %>% filter(n > 1) 
#337 dates with multiple birthdays 

mn_deaths %>% distinct(FIRSTNAME, MIDDLENAME, LASTNAME)
#3805 distinct names

mn_deaths %>% count(FIRSTNAME, MIDDLENAME, LASTNAME) %>% arrange(desc(n))
# Eric James Anderson has two entries. Lisa Marie Smith has two entries. 

mn_deaths %>% filter(FIRSTNAME == "ERIC" & MIDDLENAME == "JAMES" & LASTNAME == "ANDERSON")
# Although two entries, the names have different STATEID's and different birth dates. 

mn_deaths %>% filter(FIRSTNAME == "LISA" & MIDDLENAME == "MARIE" & LASTNAME == "SMITH")
# Although two entries, the names have different STATEID's and different birth dates. 
```

2.  check / clean up HISPANICETHNICITY (into standardized values)
```{r}
mn_deaths %>% select(HISPANICETHNICITY)

mn_deaths %>% count(HISPANICETHNICITY)

mn_deaths %>% mutate(HISPANICETHNICITY_UPPER = str_to_upper(HISPANICETHNICITY)) %>% select(HISPANICETHNICITY, HISPANICETHNICITY_UPPER)

mn_deaths %>% count(HISPANICETHNICITY_UPPER)

mn_deaths <- mn_deaths %>% mutate(HISPANICETHNICITY_UPPER = str_to_upper(HISPANICETHNICITY))
 

mn_deaths %>% mutate(HISPANICETHNICITY_PROPER = case_when(
  HISPANICETHNICITY_UPPER=="NON-HISPANIC" ~ "NOT HISPANIC",
  HISPANICETHNICITY_UPPER=="NOT-HISPANIC" ~ "NOT HISPANIC",
  HISPANICETHNICITY_UPPER=="NOT HISPANIC" ~ "NOT HISPANIC",
  HISPANICETHNICITY_UPPER=="HISPANIC" ~ "HISPANIC",
  HISPANICETHNICITY_UPPER=="UNKNOWN" ~ "UNKNOWN"
)) %>% 
  count(HISPANICETHNICITY_PROPER)

mn_deaths <- mn_deaths %>% mutate(HISPANICETHNICITY_PROPER = case_when(
  HISPANICETHNICITY_UPPER=="NON-HISPANIC" ~ "NOT HISPANIC",
  HISPANICETHNICITY_UPPER=="NOT-HISPANIC" ~ "NOT HISPANIC",
  HISPANICETHNICITY_UPPER=="NOT HISPANIC" ~ "NOT HISPANIC",
  HISPANICETHNICITY_UPPER=="HISPANIC" ~ "HISPANIC",
  HISPANICETHNICITY_UPPER=="UNKNOWN" ~ "UNKNOWN"
))
                     
```

3.  convert BIRTDATE & DEATHDATE
```{r}
mn_deaths %>% select(BIRTHDATE, DEATHDATE)

mn_deaths %>% mutate(NEWBIRTHDATE = mdy(BIRTHDATE)) %>% 
  select(NEWBIRTHDATE, BIRTHDATE)

mn_deaths %>% mutate(NEWDEATHDATE = mdy(DEATHDATE)) %>% 
  select(NEWDEATHDATE, DEATHDATE)

mn_deaths <- mn_deaths %>% mutate(NEWBIRTHDATE = mdy(BIRTHDATE), .after= BIRTHDATE) %>% 
  mutate(NEWDEATHDATE = mdy(DEATHDATE), .after=DEATHDATE)
```


TASKS: `state_pop`

1.  Reorganize the file so that you can calculate the percent change for each states population from 2018 to 2022. 
```{r}
glimpse(state_pop)

 state_pop %>% pivot_wider(names_from=year, values_from=value)

state_pop_wide <- state_pop %>% pivot_wider(names_from=year, values_from=value)

```

2.  Which state has increased the most? 
Idaho increased most in population with a 9.5% percentage change. 
```{r}
state_pop_wide %>% 
  arrange(NAME) %>% 
  mutate(pct_change = ((`2022` - `2018`)/`2022`) * 100) %>% 
  select(NAME, pct_change)

# i tried to use the tipsheet that you have us for percent change but it did not seem correct for this problem? i used the formula on this site https://www.bls.gov/cpi/factsheets/calculating-percent-changes.htm#:~:text=To%20find%20the%20percent%20change,multiply%20the%20result%20by%20100.

state_pop_wide %>% 
  arrange(NAME) %>% 
  mutate(pct_change = ((`2022` - `2018`)/`2022`) * 100) %>% 
  select(NAME, pct_change) %>%  
  arrange(desc(pct_change))
  
```

3.  Which states have decreased in population? 
According to the data, 7 states have decreased in population (Alaska, California, District of Columbia, Illinois, Louisiana, Mississippi, and West Virginia). 
```{r}
state_pop_wide %>% 
  arrange(NAME) %>% 
  mutate(pct_change = ((`2022` - `2018`)/`2022`) * 100) %>% 
  select(NAME, pct_change) %>% 
  filter(pct_change < 0)
```

4.  Where does Missouri rank? 
Missouri ranks 40th of all the states, including D.C. and Puerto Rico, regarding population percentage change having the 40th largest increase in population (.83%).
```{r}
state_pop_wide %>% 
  arrange(NAME) %>% 
  mutate(pct_change = ((`2022` - `2018`)/`2022`) * 100) %>% 
  select(NAME, pct_change) %>%  
  filter(NAME == "Missouri")

state_pop_wide %>% 
  arrange(NAME) %>% 
  mutate(pct_change = ((`2022` - `2018`)/`2022`) * 100) %>% 
  select(NAME, pct_change) %>%  
  arrange(desc(pct_change))
```

