---
title: "Campaign Finance"
output: 
---

FEC bulk data URL: https://www.fec.gov/data/browse-data/?tab=bulk-data
Data from the 2021-2022 election cycle - Midterm election

```{r}
library(tidyverse)
library(lubridate)
```

```{r}
transactions <- read_csv("data/transactions.csv")
transaction_types <- read_csv("data/transaction_types.csv")
candidates <- read_csv("data/candidates.csv")
committees <- read_csv("data/committees.csv")
```

What transaction types are we working with?
```{r}
transactions %>% 
  count(transaction_tp)
```

Join two transaction types
Test first
```{r}
transaction_types %>% 
  count(transaction_tp) %>% 
  arrange(desc(transaction_tp))
#This gives you the count, you're looking for nothing greater than one.

#OR 
transaction_types %>% 
  count(transaction_tp) %>% 
  filter(n > 1)
#No duplicates 
```

Left join
```{r}
transactions %>% 
  left_join(transaction_types, by = "transaction_tp")
#first check: are the results = rows in main table?

#second check: are there any rows where the joined columns are NA?
transactions %>% 
  left_join(transaction_types, by = "transaction_tp") %>% 
  filter (is.na(transaction_desc))
# no missing values 

# check the code and desc matchups
transactions %>% 
  left_join(transaction_types, by = "transaction_tp") %>% 
  count(transaction_tp, transaction_desc)

#optional step is to create a new table from the join: 
transaction2 <- transactions %>% 
  left_join(transaction_types, by = "transaction_tp")
```

How much money was spent for each transaction type?
```{r}
transaction2 %>% 
  group_by(transaction_desc) %>% 
  summarise(tot_amt = sum(transaction_amt)) %>% 
  arrange(desc(tot_amt))

transaction2 %>% 
  group_by(transaction_desc) %>% 
  summarise(tot_amt = sum(transaction_amt), n = n()) %>% 
  arrange(desc(tot_amt))
#n = n() is essentially the count function inside of summarise.
```

Which candidates are in our data?
```{r}
transaction2 %>% 
  count(cand_id) %>% 
  arrange(desc(n))
#35 candidates in our data

transaction2 %>% 
  count(cand_id) %>% 
  arrange(desc(n)) %>% 
  left_join(candidates, by = "cand_id")

```

Which candidate had the most money spent on them in independent expenditures?
```{r}
transaction2 %>% 
  left_join(candidates, by = "cand_id") %>% 
  filter(transaction_tp == "24A" | transaction_tp == "24E") %>% 
  group_by(cand_name) %>% 
  summarise(total_amt = sum(transaction_amt)) %>% 
  arrange(desc(total_amt))
```

```{r}
#look at opposing
transaction2 %>% 
  left_join(candidates, by = "cand_id") %>% 
  filter(transaction_tp == "24A") %>% 
  group_by(cand_name) %>% 
  summarise(total_amt = sum(transaction_amt)) %>% 
  arrange(desc(total_amt))

#look at advocating
transaction2 %>% 
  left_join(candidates, by = "cand_id") %>% 
  filter(transaction_tp == "24E") %>% 
  group_by(cand_name) %>% 
  summarise(total_amt = sum(transaction_amt)) %>% 
  arrange(desc(total_amt))
```

Can we make this code more efficient so we're not repeating ourselves
```{r}
transaction2 %>% 
  filter(transaction_tp %in% c("24A", "24E")) %>% 
  left_join(candidates, by = "cand_id") %>% 
  group_by(cand_name, transaction_tp) %>% 
  summarise(total_amt = sum(transaction_amt)) %>% 
  pivot_wider(names_from = transaction_tp, values_from = total_amt) %>% 
  mutate(total = sum(`24A`, `24E`, na.rm=T))
 
#can add percentage
transaction2 %>% 
  filter(transaction_tp %in% c("24A", "24E")) %>% 
  left_join(candidates, by = "cand_id") %>% 
  group_by(cand_name, transaction_tp) %>% 
  summarise(total_amt = sum(transaction_amt)) %>% 
  pivot_wider(names_from = transaction_tp, values_from = total_amt) %>% 
  mutate(total = sum(`24A`, `24E`, na.rm=T), opposing_pct = `24A`/total) %>% 
  arrange(desc(opposing_pct))
#mark mccloskey had 100% opposing 
```

Who are the committees spending this independent expenditure money?
```{r}
transaction2 %>% 
  filter(transaction_tp %in% c("24A", "24E")) %>% 
  left_join(committees, by = "cmte_id") %>% 
  group_by(cmte_nm) %>% 
  summarise(total = sum(transaction_amt)) %>% 
  arrange(desc(total))
```


