---
title: "Columbia Dispatch"
output: html_notebook
---

Columbia Police Dispatch:
https://www.como.gov/CMS/911dispatch/police.php
Downloaded 1/29/2024 8:43 a.m.
1/1/2019 to 1/29/2024 
This is what was requested, but we actually only got:
1/29/2023 to 1/29/2024 

```{r}
library(tidyverse)
library(lubridate)
```

Load the data
```{r}
dispatch <- read_csv("data/comodispatch.csv")
```

83,190 rows 
1 row = 1 call that resulted in a dispatch

Assessing the data
```{r}
dispatch %>% 
  group_by(CallDateTime, Address) %>% 
  summarise(n = n()) %>% 
  filter(n >1) %>% 
  arrange(desc(n))

dispatch %>% 
  filter(CallDateTime == "10/8/2023 2:08:54 PM")

#find incident number
dispatch %>% 
  group_by(InNum) %>% 
  summarise(n = n())

  #you can get the same results using the the code below
#count function finding incident number, essentially the same as group_by and summarise

  dispatch %>% 
  filter(InNum == "2023029401")
```
There are exact duplicates! - Call the city 

Let's say we learned that exact duplicates are mistakes and should be removed:
```{r}
dispatch_unique <- dispatch %>% 
  distinct()
```

Assessing every column:
```{r}
#InNum column assessment: 
dispatch_unique %>% 
  count(InNum) %>% 
  arrange(desc(n))
#still some duplicates 
```

Create a true data column for CallDateTime
```{r}
dispatch_unique %>% 
  mutate(new_date = mdy_hms(CallDateTime)) %>% 
  select(CallDateTime, new_date)

dispatch_unique <- dispatch_unique %>% 
  mutate(new_date = mdy_hms(CallDateTime))
```

What's the date range of our data?
```{r}
range(dispatch_unique$new_date)
```

Other ways to work with true dates:
```{r}
#how many dispatch calls per month
dispatch_unique %>% 
  count(month(new_date)) %>% 
  arrange(desc(n))
```







