---
title: "tidycensus"
output: html_notebook
---

```{r}
library(tidyverse)
library(tidycensus)
```

```{r}
age20 <- get_decennial(geography = "state", 
                       variables = "P13_001N", 
                       year = 2020,
                       sumfile = "dhc")
```
```{r}
mo <- get_acs(geography = "county", 
              variables = c(medincome = "B19013_001"), 
              state = "MO", 
              year = 2022)
```

```{r}
v22 <- load_variables(2022, "acs5", cache = TRUE)

v22 %>% distinct(concept) %>% 
  filter(str_detect(concept, "Plumbing"))

v22 %>% 
  filter(concept == "Tenure by Plumbing Facilities")
```

Pull all the variables from "Tenure by Plumbing Facilities" from MO
```{r}
get_acs(geography = "state", 
              variables = c(total = "B25049_001", 
                            owner_tot = "B25049_002", 
                            owner_complete= "B25049_003", 
                            owner_lacking= "B25049_004", 
                            rent_tot= "B25049_005", 
                            rent_complete= "B25049_006", 
                            renter_lacking= "B25049_007"), 
              state = "MO", 
              year = 2022)

#the nation
get_acs(geography = "state", 
              variables = c(total = "B25049_001", 
                            owner_tot = "B25049_002", 
                            owner_complete= "B25049_003", 
                            owner_lacking= "B25049_004", 
                            rent_tot= "B25049_005", 
                            rent_complete= "B25049_006", 
                            renter_lacking= "B25049_007"), 
              year = 2022)
```


*Homework 2/28*

1. Are there any counties in the US that have a median age below 30? How many? Which county has the highest median age?

There are counties in the US that have a median age below 30, with many of them with very low margin of error. There are 52 counties listed that have a median age below 30 years old. The county that has the highest median age is Sumter County, Florida with a median age of 68.3 years old (0.2 margin of error).

```{r}
v22 <- load_variables(2022, "acs5", cache = TRUE)

v22 %>% distinct(concept) %>% 
  filter(str_detect(concept, "Median Age")) 

v22 %>% 
 filter (concept == "Median Age by Place of Birth in the United States")
#Total median age by place of birth in the US: B06002_001

#seeing if there are counties with a median age lower than 30
get_acs(geography = "county", 
              variables = c(medianage = "B06002_001"),
              year = 2022) %>% 
  arrange(estimate)

#how many lower than 30?
get_acs(geography = "county", 
              variables = c(medianage = "B06002_001"),
              year = 2022) %>% 
  filter(estimate < "30")
#52 rows

# who has the highest median age
get_acs(geography = "county", 
              variables = c(medianage = "B06002_001"),
              year = 2022) %>% 
  arrange(desc(estimate))
```



2. What percentage of Missouri owner- and renter-occupied homes lack complete plumbing facilities? Are those percentages higher or lower than the national percentages? (hint: "us" is a geography)

In Missouri, 0.4% of occupied owner- and renter-occupied homes lack complete plumbing facilities. This is only slightly over the national percentage of the national percentage of occupied owner- and renter-occupied homes lacking plumbing, with the US percentage at 0.39%. 
(MO: 0.3967744, USA:0.003872237)

```{r}
get_acs(geography = "state", 
              variables = c(total = "B25049_001", 
                            owner_tot = "B25049_002", 
                            owner_complete= "B25049_003", 
                            owner_lacking= "B25049_004", 
                            rent_tot= "B25049_005", 
                            rent_complete= "B25049_006", 
                            renter_lacking= "B25049_007"), 
              state = "MO", 
              year = 2022)

#Total of occupied homes: 2458324, Total owned homes: 1661854, Total owned homes complete plumbing: 1656444, Total owned homes lacking plumbing: 5410, Renters total: 796470, Renters complete plumbing: 792156, Renters lacking plumbing: 4314) 

5410+4314 # = 9724
9754/2458324 # = 0.003967744 
# 0.4% of MO occupied homes lack plumbing


#the nation
get_acs(geography = "us", 
              variables = c(total = "B25049_001", 
                            owner_tot = "B25049_002", 
                            owner_complete= "B25049_003", 
                            owner_lacking= "B25049_004", 
                            rent_tot= "B25049_005", 
                            rent_complete= "B25049_006", 
                            renter_lacking= "B25049_007"), 
              year = 2022)
#Total of occupied homes: 125736353, Total owned homes: 81497760, Total owned homes complete plumbing: 81247774, Total owned homes lacking plumbing: 249986, Renters total: 44238593, Renters complete plumbing: 44001698, Renters lacking plumbing: 236895) 

249986+236895 # = 486881
486881/125736353 # = 0.003872237
# 0.39% of US occupied homes lack plumbing

```


3. How many households don't have internet in MO? How does that compare to national percentages? (hint: table is "Internet Subscriptions in Household")

According to the census data, 255,532 households in Missouri have no internet access. This is about 10.4% of the households in the state. The percentage of households in Missouri who do not have internet access is worse than that of the nation as a whole, with roughly 9% of all households in the United States without internet access. 

```{r}
v22 %>% 
 filter (concept == "Internet Subscriptions in Household")
# no internet access: B28011_008

get_acs(geography = "state",
        variables = c(total= "B28011_001", nointernet = "B28011_008"),
        state = "MO",
        year = 2022)
# total households in MO = 2458324, total households in MO w/o internet = 255532

255532/2458324 # = 0.1039456
# 10.4%

#nation
get_acs(geography = "us",
        variables = c(total= "B28011_001", nointernet = "B28011_008"),
        year = 2022)
# total households in US = 125736353, total households in US w/o internet = 11328375

11328375/ 125736353 # = 0.09009626
# 9%
```


4. Explore the tables in ACS 5-yr 2022 data to try to come up with an interesting sentence about Columbia or Missouri (or your home city or state).

In Missouri, 258,202 people born in in the state were not U.S. citizens at birth, therefore, foreign-born. This population makes up 4% of all Missouri births. 

(According to the census.gov, "The foreign-born population includes anyone who is not a U.S. citizen at birth, including those who become U.S. citizens through naturalization.")

```{r}
load_variables(2022, "acs5", cache = TRUE)

#I went to data.census.gov to browse through data tables to see what were some of the things store than may be interesting
v22 %>% distinct(concept) %>% 
  filter(str_detect(concept, "Foreign-Born")) 

v22 %>% 
 filter (concept == "Place of Birth for the Foreign-Born Population in the United States")
# total = B05006_001

get_acs(geography= "state",
        variable = "B05006_001",
        state = "MO",
        year = 2022)
#258202 people in MO were foreign-born (not US citizens at the time of birth). 

# how many people born in MO who are US natives
v22 %>% distinct(concept) %>% 
  filter(str_detect(concept, "Birth")) 

v22 %>% 
 filter (concept == "Place of Birth by Nativity and Citizenship Status")
# B05002_001

get_acs(geography= "state",
        variable = "B05002_001",
        state = "MO",
        year = 2022)
#6154422 people in MO were native-born (US citizens at the time of birth)

6154422+258202 # = 6412624
258202/6412624 # = 0.04026464
# 4% of the population were foreign-born

#TRYING TO FIND CITY OF COLUMBIA 

get_acs(geography= "city",
        variable = "B05006_001",
        city = "Columbia",
        year = 2022)
# this didn't work, wanted to see what Columbia would be specifically. I ended up going back to the "get_acs" explainer website and I don't think we can find cities by name but by zip code? I tried it below. (I googled Columbia MO zips)

get_acs(geography= "zcta",
        variable = "B05006_001",
        zcta = c(Columbia = "65201", 
                 "65202", "65203", 
                 "65205", "65211",
                 "65212", "65215",
                 "65216", "65217", "65218"),
        year = 2022)
#45280680 // yeah i dont think this is correct. the number is too high. 

get_acs(geography= "zcta",
        variable = "B05006_001",
        zcta = c(Columbia = "65201"),
        year = 2022)
# i can find one area code at a time, but i am not sure how to find multiple? 

#nation amount
get_acs(geography= "us",
        variable = "B05006_001",
        year = 2022)
#45280680 total people in the US were born in a foreign nation. 

```

