---
title: "R Notebook"
output: html_notebook
---

```{r}
# don't need to run this code again
install.packages("tidyverse")
```

We load libraries at the beginning of every script

```{r}
library(tidyverse)
```

Load some data

```{r}
weather <- read_csv("data/weather_events.csv")
```

Main tidyverse functions

```{r}
# SORTING FUNCTION
weather %>% arrange(desc(date))

weather %>% arrange(desc(injuries))

# FILTERING 
weather %>% filter(type== "tornado") %>% arrange(desc(fatalities))

weather %>% filter(fatalities > 0)

# operators
# ==
# !=
# >, <, >=, <=
# AND, OR: &, |

weather %>% filter(type == "tornado" & fatalities > 0)

weather %>% filter(type == "tornado" | type == "wind")

# GROUPING
weather %>% group_by(type) %>% summarise(total_injuries = sum(injuries)) %>% arrange(desc(total_injuries))
```

**Homework 1/25:**

1.  What was the most common weather event type in your home state? *or* Which weather event happened most in Illinois?

    Wind was the most common weather event in Illinois in this data set. 

```{r}
weather %>% filter(state == "IL") %>% 
  group_by (type) %>% 
  summarise(number_of_events = n()) %>%
  arrange(desc(number_of_events))
  
```

2.  How many injuries and fatalities were associated with it? *or* How many injuries did Illinois have regarding wind weather events?

    There were 659 total combined injuries and fatalities listed in the data from wind events in Illinois.

```{r}
#ATTEMPTS
# weather %>% filter(state == "IL" & type == "wind") %>% summarise(numbers_of_events = n())

# weather %>% filter(state == "IL" & type == "wind") %>%  group_by (type) %>% summarise(total_injfat = sum(injuries & fatalities))

weather %>% 
  filter(state == "IL" & type == "wind") %>% 
  group_by (type) %>% 
  summarise(total_injfat = sum(injuries, fatalities)) 


```

```{r}
#CHECK WORK
#SUM OF INJURIES IN IL FROM WIND = 628
weather %>% 
  filter(state == "IL" & type == "wind") %>% 
  group_by (type) %>% 
  summarise(total_inj = sum(injuries)) 

#SUM OF FATALITIES IN IL FROM WIND = 31
weather %>% 
  filter(state == "IL" & type == "wind") %>% 
  group_by (type) %>% 
  summarise(total_fat = sum(fatalities)) 
```

3.  Which weather event type has the highest fatality rate?

    The weather event type with the highest number of fatalities was tornado with 6,008 fatalities.

```{r}
weather %>% 
  group_by(type) %>% 
  summarise(total_fat = sum(fatalities)) %>% 
  arrange(desc(total_fat))

```

```{r}
# CHECKING WORK
weather %>% filter(type == "tornado") %>% summarise(total_fat = sum(fatalities))

weather %>% filter(type == "wind") %>% summarise(total_fat = sum(fatalities))

weather %>% filter(type == "hail") %>% summarise(total_fat = sum(fatalities))
```

4.  Which county has the most weather events? (use `cnty1_fips`)

    County 31 had the most weather events with 3,011,275,773 weather events present in the data.

```{r}
weather %>% 
  group_by (cnty1_fips) %>% 
  summarise(total = sum(event_num)) %>% 
  arrange(desc(total))
```

```{r}
#CHECK WORK
weather %>% 
  group_by (cnty1_fips) %>% 
  summarise(total = sum(event_num)) %>% 
  arrange(total)
```

5.  Are weather events increasing over time?

    Yes, weather events are increasing over time with few years averaging outside of the gradual increase.

```{r}
# ATTEMPT
# weather %>% group_by(year) %>% summarise (avg = avg(event_num))

# plot(weather %>% group_by (year) %>% summarise(avg = mean(event_num))) 

weather %>%
  group_by (year) %>% 
  summarise(avg = mean(event_num)) 

```

```{r}
#CHECK
weather %>%
  group_by (year) %>% 
  summarise(avg = mean(event_num)) %>% 
  arrange(desc(avg))
```
